% calculate number of pages from
% \label{firstcontentpage} to \label{lastcontentpage} inclusive
\makeatletter
\@tempcnta=\getpagerefnumber{lastcontentpage}\relax%
\advance\@tempcnta by -\getpagerefnumber{firstcontentpage}%
\advance\@tempcnta by 1%
\xdef\contentpages{\the\@tempcnta}%
\makeatother

\begingroup
\onehalfspacing
\section*{Declaration}

I,
\makeatletter\ifanonymised
\candidate,
\else
\authorself{}\ of \college{},
\fi\makeatother
being a candidate for \coursefor{}, hereby declare that this report and
the work described in it are my own work, unaided except as may be
specified below, and that the report does not contain material that
has already been used to any substantial extent for a comparable
purpose.
In~preparation of this report, I adhered to the
\href{https://www.cst.cam.ac.uk/files/ai_policy.pdf}{Department of
Computer Science and Technology AI Policy}. I~am content for
my report to be made available to the students and staff of the
University.

\subsection*{Contribution statement}
As part of my work on the findings described in Chapter \ref{star}, I co-authored a research paper together with my supervisors. The paper was accepted for publication in proceedings to appear later this June, and it is referenced as such. I~chose to base Chapter \ref{star} on the contents of the paper, and thus my supervisors had influence on its final form. However, all of the text in the chapter and all major technical contributions presented are my own work.
% and I declare the following about its contents: 
% \begin{itemize}
%     \item All of the text presented in this report was written directly by me.
%     \item In writing of the paper, my supervisors -- besides providing their guidance and advice on the strategy -- helped with editing, suggested references, and initiated some technical descriptions. This influenced the final form of the paper, and thus the chapter. However, I made major revisions to their descriptions when accounting for the technical details of my design. 
%     \item All of the contributions presented in the paper and this chapter are my own invention, and nearly all were present in a draft before the involvement of my supervisors. In particular, the most technically challenging part -- the proof of type safety (Appendix \ref{extra:star-proofs}) -- was produced by me.
% \end{itemize} 
I am the sole author of all other chapters.

\ifanonymised
\subsection*{Length}

\paragraph{Main chapters} 
\textbf{11988}\footnote{Computed with \texttt{texcount -inc}, applying \texttt{\%TC:ignore} outside the main-matter.} words, \textbf{\contentpages}~pages
(\pageref{firstcontentpage}--\pageref{lastcontentpage})

% \paragraph{Total} \pageref{LastPage} pages, excluding front-matter.
\fi

\vfill

\textbf{Signed:} 
\ifanonymised
\candidate
\else
\authorself{}
\fi

\vspace{0.5cm}
\textbf{Date:} \submissiondate{}
\vspace{3cm}

\endgroup
